\chapter*{Acknowledgements}
\label{ch:acknowledgements}

I would like to express my sincere gratitude to everyone who made this thesis possible.

First and foremost, I thank my thesis supervisor for their exceptional guidance, support, and valuable feedback throughout this research project. Their expertise in entrepreneurship and technology was instrumental in shaping both the technical implementation of Pista and the research methodology. I am particularly grateful for their facilitation of the comparative analysis with Winds2Ventures, which enabled the statistical validation that forms the core of this thesis. Their mentorship helped me navigate the complexities of GenAI evaluation systems and statistical analysis.

I extend my appreciation to the Winds2Ventures team for generously providing access to their evaluation platform for comparative research. Their willingness to participate in this academic study contributed significantly to the depth and validity of the analysis. This collaboration exemplifies the value of industry-academia partnerships in advancing research.

I thank the university entrepreneurship programs and competition organizers who provided access to the startup pitch datasets used in this research. Their openness to share these materials enabled the empirical validation of both evaluation systems and made the statistical comparison possible.

Special thanks to the open-source community and the developers of Next.js, Convex, OpenAI, and other technologies that made building Pista possible. Their tools and documentation enabled rapid development and deployment of a full-featured evaluation platform.

I am grateful to my professors and classmates who provided feedback during presentations and discussions that helped refine my understanding of the research questions and methodology.

Finally, I thank my family and friends for their unwavering patience, understanding, and encouragement during the research and writing process. Their support kept me motivated through challenging moments and helped maintain perspective on the broader goals of this work.

This research was conducted with integrity and academic honesty. Any remaining limitations, errors, or omissions are entirely my responsibility.
